The low-frequency Murchison radio environment:\\interference analysis and mitigation

### 2014-10-24 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Lots of updates

The \textsc{aoflagger} software provides a C++ library%TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
that can be integrated in a pipeline, thereby minimizing reading/writing of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection. Additionally, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume. It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Various tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot the statistics over various dimensions in an interactive manner.

@@

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-stddev-per-set}\vspace{-1cm}
\caption{Observed noise in the high-resolution channels for each observation. This is the residual noise in the visibilities after RFI detection. Observation `GLEAM 2014-03-17' shows DVB contamination around 180~MHz and higher, and a few channels in the FM and 2-m amateur bands show a deviating standard deviation.}
\label{fig:stddev-per-set}
\end{center}
\end{figure*}

\section{Detection results}
The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger}. This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% detections. Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI. This is mostly caused by the FM bands, which are only observed in GLEAM observations. The EoR low-band night shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%. The total RFI occupancy in all observations is 0.96\%. Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global expected RFI ratio in the 72.3--230.8 MHz range.

@@

In the ideal case, the residual visibility distributions show a smooth Rayleigh curve (see \citealt{offringa-rfi-distributions}). However, especially the tails of the residual curves of the EoR high-band and low-band observations are not smooth in Fig.~\ref{fig:lognlogs}, and the residual curve shows RFI that has not been properly detected as such. In the case of EoR low-band, samples are not detected as RFI even though they are an order of magnitude brighter than the brightest samples in the GLEAM observations. The reason such samples leak through detection is because the detection strategy includes a bandpass correction per 1.28-MHz sub-band. When RFI persistently contaminates approximately a full subband, its effect will be absorbed in the bandpass correction and becomes invisible for the flagger. In some cases, this will cause the calibration to fail on those channels, after which it will be flagged nevertheless. In other cases, such RFI is included in the imaging step or further processing, which is obviously very undesirable.

The distribution curves do not show a very significant power-law tail. Fits over regions selected by eye have been added to Fig.~\ref{fig:lognlogs}. The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N=\beta S^{-4.47}$, with $\beta$ being an arbitrary scaling factor. The GLEAM and EoR low-band observations do show a small power-law part, but the fitted exponent depends strongly on what part of the tail is selected. The selected regions result in $N=\beta S^{-1.43}$ and $N=\beta S^{-1.33}$ for the GLEAM and Eor-low curves, respectively.

@@

\section{Discussion}
\begin{itemize}
 \item SKA -- how to implement a Cotter for SKA?
 \item PAPER etc needs to use combinatorial/morphological flagging
 \item The presence of strong interference implies there might be weak interference present that is easily seen or detected. 
\end{itemize}
\subsection{Comparison with LOFAR}

>>>>

The \textsc{aoflagger} software provides a C++ library%TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
that can be integrated in a pipeline, thereby minimizing reading/writing of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection. RFI detection is only performed on cross-correlations. Auto-correlations are normally removed, because they are not used in imaging. In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume. It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Various tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot the statistics over various dimensions in an interactive manner.

@@

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-stddev-per-set}\vspace{-1cm}
\caption{Observed noise in the high-resolution channels for each observation. This is the residual noise in the visibilities after RFI detection. A few channels in the FM bands show a deviating standard deviation, and observation `GLEAM 2014-03-17' shows DVB contamination around 180~MHz.}
\label{fig:stddev-per-set}
\end{center}
\end{figure*}

\section{Detection results}
The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations. This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% detections. Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI. This is mostly caused by the FM bands, which are only observed in GLEAM observations. The EoR low-band night shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%. The total RFI occupancy in all observations is 0.96\%. Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global expected RFI ratio in the 72.3--230.8 MHz range.

@@

The distribution curves of RFI-detected samples do not show a very significant power-law tail. Fits over regions selected by eye have been added to Fig.~\ref{fig:lognlogs}. The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N=\beta S^{-4.47}$, with $\beta$ being a scaling factor. The GLEAM and EoR low-band observations do show a small power-law part, but the fitted exponent depends strongly on what part of the tail is selected. The selected regions result in $N=\beta S^{-1.43}$ and $N=\beta S^{-1.33}$ for the GLEAM and Eor-low curves, respectively.

In the ideal case, the residual visibility distributions show a smooth Rayleigh curve (see \citealt{offringa-rfi-distributions}). The residual curves in Fig.~\ref{fig:lognlogs} do not go down as quickly as a Rayleigh curve would, and therefore the distributions have a slight excess of samples with higher amplitudes. This excess is caused by the multi-variate nature of the data, for example because the noise level increases when the Galaxy goes through the beam. The EoR-high distribution has some extra features in its tails that are not smooth. Further analyses showed that these are also caused by the Galaxy, causing a few short baselines to observe samples with high amplitudes. Visual inspection of snapshots with a non-smooth residual distribution tail did not show residual RFI in such sets.

@@

\section{Discussion}
\begin{itemize}
 \item SKA -- how to implement a Cotter for SKA?
 \item PAPER etc needs to use combinatorial/morphological flagging
 \item The presence of strong interference implies there might be weak interference present that is not easily seen or detected. 
\end{itemize}

\noindent\begin{figure}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=9cm]{img/LOFAR-MWA-occupancy}
\caption{Comparison between LOFAR and MWA RFI occupancies. The statistics are regridded to the same frequency resolution of 48~kHz. }
\label{fig:lofar-mwa-occupancy}
\end{center}
\end{figure}

\subsection{Comparison with LOFAR}
The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows to compare the RFI situations at the same frequencies. The RFI occupancies for LOFAR from \citet{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}. The statistics are regridded to the same frequency resolution of 48~kHz. The average RFI occupancy in this frequency range is 1.65\% for the MWA, while O2012 reports a 3.18\% occupancy for LOFAR.

It should be noted that the RFI detection for LOFAR was performed at 0.78~kHz while for the MWA it is performed at 40~kHz. As shown in O2012, LOFAR observes RFI that is only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions. O2012 also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA. In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources: Cassiopeia~A is never visible at the MWA, and Cygnus~A is only visible at low declinations. An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations. Taking into account these differences, it becomes hard to compare the environments fairly based on detected RFI levels. Nevertheless, for both telescopes, the automatic detection strategies have been optimized such that as little data as possible is thrown away but the detection is sufficient for further data reduction. For the MWA, this goal was reached by throwing away a factor of two less data.

After RFI detection, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging. While the difference between a 1.65\% or 3.18\% loss of data is hardly a reason to spend resources to build at a remote site like the Murchison Radio Observatory, there are additional reasons to do so. A benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the FM-station bands (88--108~MHz). Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}. Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR. Such a detection requires very deep integrations, and it could be that RFI becomes problematic when reaching lower noise levels. However, initial results with LOFAR predict that RFI will not prevent a detection of EoR signals \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.
